---
title: Thermal Tolerance Landscapes of *Skistodiaptomus pallidus*
date: "`r Sys.Date()`"
output: 
  html_document:
          code_folding: hide
          code_download: true
          toc: true
          toc_float: true
  github_document:
          html_preview: false
          toc: true
          toc_depth: 3
---

```{r setup, include=T, message = F, warning = F, echo = F}
knitr::opts_chunk$set(
  echo = knitr::is_html_output(),
  fig.align = "center",
  fig.path = "../Figures/markdown/",
  dev = c("png", "pdf"),
  message = FALSE,
  warning = FALSE,
  collapse = T
)

theme_matt = function(base_size = 18,
                      dark_text = "grey20"){
  mid_text =  monochromeR::generate_palette(dark_text, "go_lighter", n_colours = 5)[2]
  light_text =  monochromeR::generate_palette(dark_text, "go_lighter", n_colours = 5)[3]
  
  ggpubr::theme_pubr(base_family="sans") %+replace% 
    theme(
      panel.background  = element_rect(fill="transparent", colour=NA), 
      plot.background = element_rect(fill="transparent", colour=NA), 
      legend.background = element_rect(fill="transparent", colour=NA),
      legend.key = element_rect(fill="transparent", colour=NA),
      text = element_text(colour = mid_text, lineheight = 1.1),
      title = element_text(size = base_size * 1.5,
                           colour = dark_text),
      axis.text = element_text(size = base_size,
                               colour = mid_text),
      axis.title.x = element_text(size = base_size * 1.2,
                                  margin = unit(c(3, 0, 0, 0), "mm")),
      axis.title.y = element_text(size = base_size * 1.2,
                                  margin = unit(c(0, 5, 0, 0), "mm"), 
                                  angle = 90),
      legend.text = element_text(size=base_size * 0.9),
      legend.title = element_text(size = base_size * 0.9, 
                                  face = "bold"),
      plot.margin = margin(0.25, 0.25, 0.25, 0.25,"cm")
    )
}

theme_matt_facets = function(base_size = 18,
                             dark_text = "grey20"){
  mid_text =  monochromeR::generate_palette(dark_text, "go_lighter", n_colours = 5)[2]
  light_text =  monochromeR::generate_palette(dark_text, "go_lighter", n_colours = 5)[3]
  
  theme_bw(base_family="sans") %+replace% 
    theme(
      panel.grid = element_blank(),
      panel.background  = element_rect(fill="transparent", colour=NA), 
      plot.background = element_rect(fill="transparent", colour=NA), 
      legend.background = element_rect(fill="transparent", colour=NA),
      legend.key = element_rect(fill="transparent", colour=NA),
      text = element_text(colour = mid_text, lineheight = 1.1),
      strip.text.x = element_text(size = base_size),
      title = element_text(size = base_size * 1.5,
                           colour = dark_text),
      axis.text = element_text(size = base_size,
                               colour = mid_text),
      axis.title.x = element_text(size = base_size * 1.2,
                                  margin = unit(c(3, 0, 0, 0), "mm")),
      axis.title.y = element_text(size = base_size * 1.2,
                                  margin = unit(c(0, 5, 0, 0), "mm"), 
                                  angle = 90),
      legend.text = element_text(size=base_size * 0.9),
      legend.title = element_text(size = base_size * 0.9, 
                                  face = "bold"),
      plot.margin = margin(0.25, 0.25, 0.25, 0.25,"cm")
    )
}
```

```{r}

raw_data |> 
  select(set_temp_c, read_temp_c) |> 
  distinct() |> 
  ggplot(aes(x = set_temp_c, y = read_temp_c)) + 
  geom_abline(intercept = 0, slope = 1) + 
  geom_point() + 
  theme_matt()

```

```{r}

raw_data |> group_by(source_file) |> 
  count(read_temp_c) |> 
  ggplot(aes(x = read_temp_c, y = n)) + 
  geom_bar(stat = "identity")

```

## Building the survival dataset

`responding == "no"` indicates death. Rows flagged for data-quality issues (`flagged == "yes"`) are excluded before constructing one time-to-event record per individual: the death time is the first timepoint at which an individual is recorded as not responding, or, if it never stops responding, the last observed timepoint (right-censored). The actual measured bath temperature (`read_temp_c`) is used as the temperature covariate rather than the nominal `set_temp_c`, since the two diverge noticeably at higher target temperatures.

```{r build-survival-data}

surv_data = raw_data |>
  filter(flagged != "yes" | is.na(flagged)) |>
  arrange(source_file, water_bath_id, stopwatch_start_time, well_id, timepoint_min) |>
  group_by(source_file, water_bath_id, stopwatch_start_time, well_id, population, read_temp_c) |>
  summarise(
    time = if (any(responding == "no")) min(timepoint_min[responding == "no"]) else max(timepoint_min),
    event = as.integer(any(responding == "no")),
    n_obs = n(),
    .groups = "drop"
  )

surv_data
```

## All data 

Individual survival times are modeled jointly across all temperatures using an accelerated failure time (AFT) model, rather than fitting separate Kaplan-Meier curves per temperature, since many temperature/population cells have only 10 individuals. Weibull, log-normal, and log-logistic distributions are compared by AIC.

```{r compare-aft-distributions}
dists = c("weibull", "lognormal", "loglogistic")

aic_table = map_dfr(dists, function(d) {
  m = survreg(Surv(time, event) ~ population * read_temp_c, data = surv_data, dist = d)
  tibble(dist = d, AIC = AIC(m), logLik = as.numeric(logLik(m)))
})

aic_table
```

The log-logistic distribution has the lowest AIC. A likelihood ratio test comparing models with and without a population × temperature interaction shows the interaction is not supported over the full temperature range, so the additive model is used here.

```{r fit-full-aft-model}

full_model_interaction = survreg(Surv(time, event) ~ population * read_temp_c, data = surv_data, dist = "loglogistic")

full_model_tidy = tidy(full_model_interaction, conf.int = TRUE)
full_model_tidy
```

Each 1°C increase in read temperature is associated with median survival time shrinking to about `r scales::percent(exp(full_model_tidy$estimate[full_model_tidy$term == "read_temp_c"]))` of its prior value (time ratio `r round(exp(full_model_tidy$estimate[full_model_tidy$term == "read_temp_c"]), 4)`, p `r format.pval(full_model_tidy$p.value[full_model_tidy$term == "read_temp_c"], digits = 2)`) — a steep, well-supported decline over the 31-34.8°C range tested. There is no evidence of a difference between populations once temperature is accounted for (time ratio `r round(exp(full_model_tidy$estimate[full_model_tidy$term == "populationOPOH"]), 2)`, p `r format.pval(full_model_tidy$p.value[full_model_tidy$term == "populationOPOH"], digits = 2)`), and the wide confidence interval on that estimate reflects the modest sample size.

Median survival time predicted from the additive model, by population and temperature, plotted on a log scale. Predictions are shown across the full observed temperature range (31-34.8°C); extrapolation toward the lower end is poorly constrained because almost no deaths were observed at 31°C and 33.1°C within the observation window, so these predictions should be treated with caution.

```{r plot-full-aft-predictions}
full_pred_grid = expand_grid(
  population = unique(surv_data$population),
  read_temp_c = seq(min(surv_data$read_temp_c), max(surv_data$read_temp_c), length.out = 100)
)

full_pred = predict(full_model_interaction, newdata = full_pred_grid, type = "quantile", p = 0.5, se.fit = TRUE)

full_pred_grid = full_pred_grid |>
  mutate(
    median_time = full_pred$fit,
    se = full_pred$se.fit,
    lwr = exp(log(median_time) - 1.96 * se / median_time),
    upr = exp(log(median_time) + 1.96 * se / median_time)
  )

ggplot(full_pred_grid, aes(x = read_temp_c, y = median_time, color = population, fill = population)) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.15, color = NA) +
  geom_line() +
  scale_y_log10() +
  labs(
    x = "Water bath temperature (read, °C)",
    y = "Predicted median survival time (min, log scale)",
    title = "AFT model (log-logistic): median survival time vs. temperature"
  ) +
  theme_matt()
```

## Just temperatures above 33°C

Restricting to individuals held at read temperatures above 33°C, an AFT model is fit with a population × temperature interaction, per the analysis plan, even though the interaction was not significant over the full temperature range.

```{r fit-interaction-aft-model}
surv_data_high = surv_data |> filter(read_temp_c > 33)

high_temp_model = survreg(Surv(time, event) ~ population * read_temp_c, data = surv_data_high, dist = "loglogistic")

high_temp_tidy = tidy(high_temp_model, conf.int = TRUE)
high_temp_tidy
```

The interaction term remains non-significant (p `r format.pval(high_temp_tidy$p.value[high_temp_tidy$term == "populationOPOH:read_temp_c"], digits = 2)`), so there is still no evidence that the two populations decline at different rates as temperature rises above 33°C — both show a similarly steep drop in median survival time (main effect of read temperature: time ratio `r round(exp(high_temp_tidy$estimate[high_temp_tidy$term == "read_temp_c"]), 4)` per 1°C, p `r format.pval(high_temp_tidy$p.value[high_temp_tidy$term == "read_temp_c"], digits = 2)`). The interaction is retained here per the analysis plan, but given its lack of significance and the small per-cell sample sizes, the near-parallel curves in the plot below should be interpreted cautiously rather than as confirmed evidence of population-specific slopes.

```{r plot-restricted-aft-predictions}
high_pred_grid = expand_grid(
  population = unique(surv_data_high$population),
  read_temp_c = seq(min(surv_data_high$read_temp_c), max(surv_data_high$read_temp_c), length.out = 100)
)

high_pred = predict(high_temp_model, newdata = high_pred_grid, type = "quantile", p = 0.5, se.fit = TRUE)

high_pred_grid = high_pred_grid |>
  mutate(
    median_time = high_pred$fit,
    se = high_pred$se.fit,
    lwr = exp(log(median_time) - 1.96 * se / median_time),
    upr = exp(log(median_time) + 1.96 * se / median_time)
  )

ggplot(high_pred_grid, aes(x = read_temp_c, y = median_time, color = population, fill = population)) +
  #geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.15, color = NA) +
  geom_line() +
  #scale_y_log10() +
  labs(
    x = "Water bath temperature (read, °C)",
    y = "Predicted median survival time (min, log scale)",
    title = "AFT model (log-logistic, population × temperature): median survival vs. temperature",
    subtitle = "Restricted to read temperatures above 33°C"
  ) +
  theme_matt()
```

## Tolerance Landscape Parameters

Below, we re-fit a model to each population separately, and extract the relevant model parameters (intercept and slope term).

```{r}

TTL_params = surv_data_high %>%
  group_by(population) %>%
  do(tidy(survreg(Surv(time, event) ~ read_temp_c, data = ., dist = "loglogistic"))) |> 
  mutate(term = if_else(term == "(Intercept)", "Beta_0", term)) |> 
  filter(term != "Log(scale)") |> 
  pivot_wider(id_cols = population, 
              names_from = term, 
              values_from = c(estimate, std.error)) |>  
  select(population, "Beta_0" = estimate_Beta_0, "temp_slope" = estimate_read_temp_c, 
         "std_err_beta_0" = std.error_Beta_0, "std_err_slope" = std.error_read_temp_c) |> 
  mutate(ctmax_static = -1*(Beta_0/temp_slope),
         z_slope = -1/temp_slope)

```



### Dynamic Landscapes 

Rezende et al. 2020 provide functions to 1) estimate the tolerance landscape from LT50 times and temperatures, and 2) use this landscape to predict mortality under fluctuating temperatures. Their original approach (`tolerance.landscape()`) first collapses each temperature to a single LT50 point estimate, then fits `ctmax`/`z` via lm across those points and builds the reference survival curve by interpolating raw per-temperature survival curves.

**Superseded exploratory step, not used downstream.** The chunk below (not evaluated) shows what that LT50-estimation step looks like; it is kept only for reference and is **not** the method used to build the landscape in this report. The `lt50_values` object it would produce is unused elsewhere.

```{r, eval=FALSE}

lt50_values = surv_data_high %>%
  group_by(population, read_temp_c) %>%
  nest() %>%
  mutate(
    model = map(data, ~ survreg(Surv(time, event) ~ 1, data = .x, dist = "loglogistic")),
    lt50 = map_dbl(model, ~ predict(.x, type = "quantile", p = 0.5)[1])
  ) %>%
  dplyr::select(population, lt50)

ggplot(lt50_values, aes(x = read_temp_c, y = log(lt50), colour = population)) + 
  geom_point() + 
  geom_smooth(method = "lm")

```

Instead, we fit an AFT model to the full individual-level, censored survival data for each population (as in `TTL_params`) and extract the tolerance landscape directly from the model with `model_ttl()`. This uses all individuals rather than collapsing each temperature to a single LT50 estimate first, and propagates coefficient uncertainty into `ctmax`, `z`, and the reference survival curve via simulation from `vcov(model)`.

```{r}
co_data = surv_data_high |> filter(population == "CPCO")
oh_data = surv_data_high |> filter(population == "OPOH")

co_model = survreg(Surv(time, event) ~ read_temp_c, data = co_data, dist = "loglogistic")
oh_model = survreg(Surv(time, event) ~ read_temp_c, data = oh_data, dist = "loglogistic")

co_tl = model_ttl(co_model, "read_temp_c", seed = 6053)
oh_tl = model_ttl(oh_model, "read_temp_c", seed = 7481)

tibble(
  population = c("CPCO", "OPOH"),
  ctmax = c(co_tl$ctmax, oh_tl$ctmax),
  ctmax_lwr = c(co_tl$ctmax.ci[1], oh_tl$ctmax.ci[1]),
  ctmax_upr = c(co_tl$ctmax.ci[2], oh_tl$ctmax.ci[2]),
  z = c(co_tl$z, oh_tl$z),
  z_lwr = c(co_tl$z.ci[1], oh_tl$z.ci[1]),
  z_upr = c(co_tl$z.ci[2], oh_tl$z.ci[2])
)
```

In this first example, we simulate a simple fluctuating environment. Note, the function assumes temperature for each minute is provided. As we can see, even small differences in the tolerance landscape can lead to significant differences in the observed mortality. 

```{r}
variable.temp <- rep(c(30, 34, 34.5),each=10, times = 4)

co_dl <- dynamic.landscape(variable.temp,co_tl)	
oh_dl <- dynamic.landscape(variable.temp,oh_tl)	

surv_comp = bind_rows(data.frame("time" = co_dl$time, "alive" = co_dl$alive, "pop" = "CO"),
                      data.frame("time" = oh_dl$time, "alive" = oh_dl$alive, "pop" = "OH"))

ggplot(surv_comp, aes(x = time, y = alive, colour = pop)) + 
  geom_line(linewidth = 2) + 
  theme_matt()
```

```{r}

dyn_surv = dynamic.landscape(env_temps_minutes$temp_c, oh_tl)

```

The plot below overlays the temperature and survival data to highlight the alignment between peak temperatures and drops in survival. 

```{r}

dyn_df = data.frame("time" = dyn_surv$time, 
           "temp" = dyn_surv$ta, 
           "surv" = dyn_surv$alive)

dyn_df_thin <- dyn_df |>
  filter(surv != lag(surv, default = -Inf) | row_number() == n())

temp_range <- range(dyn_df$temp)
surv_range <- range(dyn_df$surv)

surv_range_fixed <- c(0, 100)

scale_surv <- function(x) (x - surv_range_fixed[1]) / diff(surv_range_fixed) * diff(temp_range) + temp_range[1]

unscale_surv <- function(x) (x - temp_range[1]) / diff(temp_range) * diff(surv_range_fixed) + surv_range_fixed[1]

ggplot(dyn_df, aes(x = time)) +
  geom_line(aes(y = temp, color = "Temperature (°C)"), 
            linewidth = 1) +
  geom_line(data = dyn_df_thin, 
            aes(y = scale_surv(surv), color = "Survival (%)"), 
            linewidth = 2) +
  scale_y_continuous(
    name = "Temperature (°C)",
    sec.axis = sec_axis(~ unscale_surv(.), name = "Survival (%)")
  ) +
  scale_color_manual(values = c("Survival (%)" = "steelblue", "Temperature (°C)" = "firebrick")) +
  labs(x = "Time", color = NULL) + 
  theme_matt() + 
  theme(legend.position = "right")

```



